---
title: "Introduction: methodologies and implementation"
date: today
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Introduction: methodologies and implementation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `comprehensivecohort` package provides tools for analyzing data from comprehensive cohort studies in the presence of unmeasured confounding and missing outcome (either continuous or binary). The goal is to estimate the comprehensive cohort causal effects (CCCE), the average causal effect for the entire cohort. This vignette has two parts: 1). Methodologies. This part will cover details on what is Comprehensive Cohort Study design, the causal estimands, assumptions, and brief descriptions of the estimation process. 2). Implementation. This part will focus on how to use the R functions to implement the estimation of CCCE under pre-specified sensitivity parameters. We will also provide codes for presenting results. 

# Methodologies

## Comprehensive Cohort Study design

The Comprehensive Cohort Study (CCS) design is a nested study design where individuals can opt to participate in an randomized control trial (RCT) or the parallel observational study (OBS). In RCT, individuals are randomized to treatments. However, in OBS, individuals can choose the treatment options provided in RCT. All individuals are followed-up. Thus, baseline covariates, treatment and outcome information are collected for the entire cohort. 

## Causal estimands

Let $X$ be baseline covariates, $R$ randomization consent indicator ($1$ for RCT, $0$ for OBS), $T$ binary treatment indicator, $Y$ as outcome and $M$ the missing outcome indicator ($M=1$ if $Y$ is observed, $M=0$ if $Y$ is missing). Let $O = (X, R, T, Y)$ be the full data without outcome missingness, and $\widetilde{O} = (X,R,T,M,Y:M=1)$ the observed data. Let $P$ and $\widetilde{P}$ be the distribution of $O$ and $\widetilde{O}$. 

We further introduce the following notations: 

* $\widetilde{F}_{t, r}(y|x)=P(Y\leq y\mid T=t, R=r, X=x, M=1)$
* $\widetilde{\mu}_{t,r}(h(Y); x)=E[h(Y) \mid T=t, R=r, X=x, M=1]$
* $\pi_{t,r}(x)=P(T = t \mid R=r, X=x)$
* $g_r(x)=P(R = r \mid X=x)$
* $\eta_m(x, r, t)=P(M =m \mid X=x, R=r, T=t)$

Our causal estimand of interest is the **comprehensive cohort causal effect** (CCCE): $E[Y(1)]-E[Y(0)]$. We will draw inference about $E[Y(t)]$. 

## Assumptions

* Consistency: $Y(t) = Y$ if $T=t$, for $t=0,1$. 
* Positivity: $0 < P(T=1 \mid R=1) < 1$ and $0 < P(T=1 \mid R=0, X=x) < 1$ for all $x$ in the support of $X$. 
* Randomization in RCT: $T \perp \{ Y(t), X \} \mid R=1$, for $t = 0, 1$. 
* Confounding Controlled by $\gamma_t$ in OBS: 
$$dF(y(t) \mid T=1-t,R=0,X)  =  dF(y(t) \mid  T=t,R=0,X) \times  \frac{ \exp\{  \gamma_t s_t(y(t)) \} }{ \mu_{t, 0} (\exp\{  \gamma_t s_t(Y(t)) \}; X)}$$
* Missing-At-Random Assumption for Outcome: $M \perp Y \mid  T, X, R$. 

Under the assumptions, we can identify $E[Y(t)]$ as

\begin{aligned}
\psi_t(\widetilde{P}; \gamma_t) =& \underbrace{E[ \widetilde{\mu}_{t,1}(Y;X) \big| R=1]}_{\psi_{t,1}(\widetilde{P})} P(R=1)\\
&+  \underbrace{E \bigg[  \widetilde{\mu}_{t,0}(Y;X) \pi_{t,0}(X) +\frac{ \widetilde{\mu}_{t,0}(Y \! \exp\{  \gamma_t s_t(Y)\} ;X) }{\widetilde{\mu}_{t,0}( \exp\{  \gamma_t s_t(Y)\} ;X) } \pi_{1-t,0}(X) \bigg| R=0\bigg]}_{\psi_{t,0}(\widetilde{P}; \gamma_t)} P(R=0)
\end{aligned}

$\psi_{t, 1}(\widetilde{P})$ and $\psi_{t, 0}(\widetilde{P}; \gamma_t)$ are the identifications of $E[Y(t)|R=1]$ and $E[Y(t)|R=0]$. Details of estimations for $E[Y(t)|R=1]$ and $E[Y(t)|R=0]$ can be found in the paper. We define the **randomized trial causal effect** (RTCE) as $\mbox{RTCE}=\psi_{1, 1}(\widetilde{P})-\psi_{0, 1}(\widetilde{P})$; and the **patient preference causal effect** (PPCE) as $\mbox{PPCE}(\gamma_1,\gamma_0) = \psi_{1, 0}(\widetilde{P}; \gamma_1)-\psi_{0, 0}(\widetilde{P}; \gamma_0)$. 

## Methods

We will describe the main estimation steps for $\psi_{t}(\widetilde{P}; \gamma_t)$: (1) Derive the efficient influence function. (2) Estimate $\widetilde{P}$. (3) Compute the one-step, truncated, split sample estimator. (4) Compute the variance and $95\%$ confidence interval.

(1). Derive the efficient influence function for $\psi_t(\widetilde{P}; \gamma_t)$, denoted $\phi_t(\widetilde{P}; \gamma_t)(\widetilde{O})$

$$\phi_t(\widetilde{P}; \gamma_t)(\widetilde{O})
    =  \upsilon_{t, 1}(\widetilde{P})(\widetilde{O})\times P(R=1)+\upsilon_{t, 0}(\widetilde{P}; \gamma_t)(\widetilde{O})\times P(R=0) - \psi_t(\widetilde{P}; \gamma_t) \ , $$

where the exact form of $\upsilon_{t, 1}(\widetilde{P})(\widetilde{O})$ and $\upsilon_{t, 0}(\widetilde{P}; \gamma_t)(\widetilde{O})$ can be found in the paper. 


(2). Estimate $\widetilde{P}$

To construct a estimator for $E[Y(t)]$, we need to estimate $\pi_{t, r}(X)$, $\eta_1(X, r, t)$ and $\widetilde{F}_{t, r}(y|X)$ for $r=0, 1$. The estimation methods we apply have fast enough convergence rate so that our resulting estimators converge to a normal distribution at $\sqrt{n}$-rates. Specifically, we apply the generalized additive model (GAM) with a logistic link function [@hastie1990tibshirani] for $\pi_{t, r}(X)$ and $\eta_1(X, r, t)$ for $r=0, 1$. We apply the single index model [@chiang2012new] for $\widetilde{F}_{t, r}(y|X)$ for $r=0, 1$. We then estimate  $\widetilde{\mu}_{t, 0}\big(Y\exp(\gamma_t s_t( Y)); X\big)$, $\widetilde{\mu}_{t, 0}\big(\exp(\gamma_t s_t(Y)); X\big)$ and $\widetilde{\mu}_{t, r}\big(Y; X\big)$ by numeric integration. 

(3). Compute the one-step, truncated, split sample estimator 

We apply K-fold sample splitting and tuning-free Huberization procedure [@tuning_wang_2021] to truncate the k-th split $\upsilon_{t, 1} \left(\widehat{\widetilde{P}}^{(-k)}\right) (\widetilde{O}_i)$ and $\upsilon_{t, 0}\left(\widehat{\widetilde{P}}^{(-k)} ; \gamma_t\right) (\widetilde{O}_i)$: 

\begin{equation*}
    	\widehat{\psi}^\dagger_t(\gamma_t) = \frac{1}{K} \sum_{k=1}^K \underbrace{\left\{\frac{1}{n_k} \sum_{i: S_i=k}\Bigg[ \underbrace{\upsilon_{t, 1}^\dagger \left(\widehat{\widetilde{P}}^{(-k)}\right) (\widetilde{O}_i)\times P_{n/n_k}^{(-k)}(R=1) +\upsilon_{t, 0}^\dagger \left(\widehat{\widetilde{P}}^{(-k)} ; \gamma_t\right) (\widetilde{O}_i)\times P_{n/n_k}^{(-k)}(R=0)}_{\upsilon_t^\dagger\left(\widehat{\widetilde{P}}^{(-k)} ; \gamma_t\right)(\widetilde{O}_i)}\Bigg]\right\}}_{\widehat{\psi}_t^{(k)\dagger}(\gamma_t)}
\end{equation*}
where $S_i$ is the split membership of the $i$th observation (i.e, $S_i \in \{1, \ldots, K\}$). 

(4). Compute variance and $95\%$ confidence interval

We estimate the variance of $\widehat{\psi}_t^\dagger(\gamma_t)$ by
$$\sigma_t^{\dagger 2}=\frac{1}{nK}\sum_{k=1}^K\left\{\frac{1}{n_k-1}\sum_{i: S_i=k}\left\{\upsilon_t^\dagger\left(\widehat{\widetilde{P}}^{(-k)} ; \gamma_t\right)(\widetilde{O}_i)-\widehat{\psi}_t^{(k)\dagger}(\gamma_t)\right\}^2\right\}$$

The $95\%$ Wald confidence interval for $\psi_t(\widetilde{P}; \gamma_t)$ is $\widehat{\psi}^\dagger_t(\gamma_t)\pm z_{0.975}\frac{\sigma_t^\dagger}{\sqrt{n}}$. 

# Implementation

We will demonstrate how to use the R functions in `comprehensivecohort` to estimate CCCE, using an example data included in the package. First, install package directly from git: 

```{r message=FALSE}
remotes::install_github("UofUEpiBio/Comprehensive_cohort")
```

## Data

The dataset included in the package is a simulated data based on the TOIB study [@underwood2008topical]. The TOIB study is a comprehensive cohort study on the effect of topical versus oral non-steroidal anti-inflammatory drugs (NSAIDs) in managing knee pain among older adults with chronic knee pain. In the analysis, the outcome of interest is Western Ontario and McMaster Universities Osteoarthritis Index (WOMAC) pain score ($0-100$) at 12 months. We include age, baseline WOMAC pain score, expected pain a year from now and chronic pain grade as baseline covariates. 

```{r}
library(comprehensivecohort)
## load data
data(ccohort)
## data structure
str(ccohort)
```

The dataset has 563 observations and 8 variables, which include:

* Y: WOMAC pain score at 12 months, continuous variable ranging from 0 to 100. NA indicate missing outcome. 
* M: outcome missingness indicator, binary variable. $1$ if $Y$ is observed, $0$ if $Y$ is missing. 
* R: randomization consent indicator, binary variable. $1$ for RCT, $0$ for OBS. 
* t: treatment indicator, binary variable. $1$ for topical NSAIDs treatment group, $0$ for oral NSAIDs treatment group. 
* age: age, continuous variable. 
* womac_bq: WOMAC pain score at baseline, continuous variable ranging from 0 to 100. No missing value. 
* expectationb: expected pain a year from now, categorical variable. Three levels: much/a little worse, about the same, a little/much better/free of pain. 
* ChronicPainb: chronic pain grade, categorical variable. Two levels: 1-2, 3-4. 

## Application 

The main function is `est_psi()`, which outputs point estimates, variance and $95\%$ Walds confidence interval for $\psi_t(\widetilde{P}; \gamma_t)$, $\psi_{t, 0}(\widetilde{P}; \gamma_t)$ and $\psi_{t, 1}(\widetilde{P})$ under one or more pre-specified $\gamma_t$. 

Users need to input data and specify several parameters: 

* Data: <span style="color: blue;">Y</span>, <span style="color: blue;">M</span>, <span style="color: blue;">R</span>, <span style="color: blue;">t</span>, <span style="color: blue;">M</span> (baseline covariates in the form of data frame)
* Estimand of interest: <span style="color: blue;">trt</span>=1 if estimating $E[Y(1)]$, $E[Y(1)|R=0]$ and $E[Y(1)|R=1]$; <span style="color: blue;">trt</span>=0 if estimating $E[Y(0)]$, $E[Y(0)|R=0]$ and $E[Y(0)|R=1]$
* Sensitivity parameters: <span style="color: blue;">gamma</span>, a vector of $\gamma_t$ value. 
* Single index model settings [@redd2025sensiatrpackageconducting]: <span style="color: blue;">kernel</span>, kernel smoothing, choices of Gaussian and Epanechnikov kernel; <span style="color: blue;">single_index_method</span>, types of constraints in estimation, choices of setting first coefficient to 1, norm of coefficients to 1, and bandwidth to 1; <span style="color: blue;">method</span>, optimization methods, default to `optim`; <span style="color: blue;">use_mave</span>, whether to apply MAVE [@mave_xia_2002; @mave_wang_2008] or cumulative sliced inverse regression method [@slice_zhu_2010] to estimate initial value of the coeffficients, default to TRUE. 
* Truncation methods: <span style="color: blue;">simple_trunc</span>=TRUE to apply quantile truncation of $\frac{1}{\pi_{t, r}(X)}$ and $\frac{1}{\eta_m(X, r, t)}$, <span style="color: blue;">simple_trunc</span>=FALSE to apply tuning-free Huberization procedure [@tuning_wang_2021] to influence functions. If <span style="color: blue;">simple_trunc</span>=TRUE, specify quantile truncation by setting <span style="color: blue;">quant</span> from 0 to 1. 
* K-fold sample splitting: <span style="color: blue;">fold</span>=K, a integer. <span style="color: blue;">seed</span>, set.seed(seed). 

```{r}
## data adaptive truncation
out <- with(ccohort, {
  est_psi(Y, M, R, X = data.frame(age, womac_bq, expectationb, ChronicPainb), 
          t, trt = 1, gamma = c(0, 0.5), fold = 5, seed = 1, IF_output = FALSE,
          simple_trunc = FALSE, quant = NULL, kernel="dnorm", 
          single_index_method="norm1coef", method="optim")
})
```

## Results


# References








